% Three Independent Capabilities of Reflective Computation
% LICS submission — ACM SIGCONF format, double-blind
\documentclass[sigconf,screen,review,anonymous]{acmart}

% Lean code listings
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily\small,
  columns=flexible,
  breaklines=true,
  frame=none,
  xleftmargin=1em,
}

% Theorems
\newtheorem{observation}{Observation}
\newtheorem{remark}{Remark}

% Shortcuts
\newcommand{\dotop}{\cdot}
\newcommand{\FRM}{\textsf{FRM}}
\newcommand{\ICP}{\textsf{ICP}}
\newcommand{\Fin}{\mathrm{Fin}}
\newcommand{\core}{\mathrm{core}}
\newcommand{\Nat}{\mathbb{N}}

\begin{document}

\title{Three Independent Capabilities of Reflective Computation in Finite Algebras}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003790.10003795</concept_id>
<concept_desc>Theory of computation~Algebraic language theory</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10003790.10002990</concept_id>
<concept_desc>Theory of computation~Logic and verification</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Algebraic language theory}
\ccsdesc[500]{Theory of computation~Logic and verification}

\keywords{finite algebra, reflective computation, independence, Lean~4, machine-checked proofs}

\begin{abstract}
We study three algebraic properties of finite extensional 2-pointed magmas associated with reflective computation: self-representation, self-description, and self-execution. Self-representation is the existence of a section-retraction pair on the core. Self-description is the classifier dichotomy: every core element acts on the core either entirely through the distinguished absorber values $\{z_1,z_2\}$ or entirely outside them. Self-execution is the Internal Composition Property: some core action factors non-trivially through two others, one core-preserving.

Our main result is that these three capabilities are pairwise independent: no capability implies any other. The six non-implications are witnessed by Lean-verified finite counterexamples of sizes $N\!=\!5, 6, 8,$ and $10$. We also prove that the decomposition induced by self-description is invariant under isomorphism, and that associativity is incompatible with the classifier structure used for self-description. All results are formalized in Lean~4 without \texttt{sorry}.
\end{abstract}

\maketitle

% ══════════════════════════════════════════════════════════════
\section{Introduction}
% ══════════════════════════════════════════════════════════════

Meta-circular evaluators combine mechanisms for internal representation, internal classification, and internal reduction. These co-occur in practice, but we prove they are independent: no mechanism implies any other in finite extensional magmas. The connection to reflective computation is motivational: we isolate algebraic properties inspired by self-representation, self-description, and self-execution, and prove they are independent. Whether this independence lifts to richer computational settings is the primary open question (Section~\ref{sec:discussion}).

We define three capabilities of a finite extensional 2-pointed magma (a set with a binary operation, exactly two left-absorbers, and extensionality; the framework follows universal algebra~\cite{burris81}):

\begin{itemize}
  \item \textbf{Self-representation (S):} a section-retraction pair on the core, providing a basic internal coding mechanism.
  \item \textbf{Self-description (D):} the classifier dichotomy — every core element acts on the core either entirely through the absorber values $\{z_1, z_2\}$ or entirely outside them, creating a partition into zeros, classifiers, and non-classifiers.
  \item \textbf{Self-execution (H):} the Internal Composition Property (ICP) — some core action factors non-trivially as $a \dotop x = c \dotop (b \dotop x)$ with $b$ core-preserving, $a, b, c$ pairwise distinct, and $a$ taking $\geq 2$ distinct values.
\end{itemize}

The three capabilities are conceptually related but mathematically independent. In naturally structured reflective systems they often co-occur, but finite extensional 2-pointed magmas separate them completely: six Lean-verified counterexamples establish all six pairwise non-implications. A Lean-verified witness shows all three coexist at $N\!=\!6$.

The three-category decomposition induced by D (zeros, classifiers, non-classifiers) is proved to be an algebraic invariant: it is preserved by all magma isomorphisms that respect the absorbers. The ICP is proved logically equivalent to the standard operational axioms (Compose + Inert), giving self-execution a single-concept characterization on par with S and D.

All results are formalized in Lean~4~\cite{moura21} and verified by \texttt{lake build}, zero \texttt{sorry}. The proofs divide into universal algebraic theorems (proved by pure equational reasoning, no \texttt{decide}) and model-specific verifications (proved by \texttt{native\_decide} on concrete Cayley tables).

\paragraph{Contributions.}
\begin{enumerate}
  \item A characterization of H: the ICP is logically equivalent to Compose+Inert (Section~\ref{sec:icp}).
  \item An independence theorem: S, D, and H are pairwise independent, with Lean-verified counterexamples in all six directions at $N\!=\!5$ through $N\!=\!10$ (Section~\ref{sec:independence}), and a coexistence witness at $N\!=\!6$ (Section~\ref{sec:bound}).
  \item Universal algebraic obstructions: no extensional 2-pointed magma carrying both a classifier and a retraction pair has a right identity or is associative, ruling out semigroups, monoids, and groups (Section~\ref{sec:universal}).
  \item A decomposition invariance theorem: the three-category partition is preserved by isomorphisms (Section~\ref{sec:invariance}).
\end{enumerate}

% ══════════════════════════════════════════════════════════════
\section{Definitions}
\label{sec:definitions}
% ══════════════════════════════════════════════════════════════

Throughout, $(S, \dotop)$ is a finite magma on $\Fin(n) = \{0, 1, \ldots, n-1\}$ with binary operation $\dotop : \Fin(n) \times \Fin(n) \to \Fin(n)$.

\begin{definition}[Extensional 2-Pointed Magma]
\label{def:epm}
An \emph{extensional 2-pointed magma} is a tuple $(S, \dotop, z_1, z_2)$ where:
\begin{enumerate}
  \item $z_1, z_2 \in S$ are distinct \emph{left-absorbers}: $z_i \dotop x = z_i$ for all $x$.
  \item No other element is a left-absorber.
  \item \emph{Extensionality}: if $a \dotop x = b \dotop x$ for all $x$, then $a = b$.
\end{enumerate}
We write $\core(S) := S \setminus \{z_1, z_2\}$ for the non-absorber elements.
\end{definition}

\begin{definition}[Faithful Retract Magma]
\label{def:frm}
A \emph{Faithful Retract Magma} $(\FRM)$ extends an extensional 2-pointed magma with a \emph{retraction pair} $s, r \in \core(S)$ satisfying $r \dotop (s \dotop x) = x$ for all $x \in \core(S)$, with $r \dotop z_1 = z_1$ (anchoring). The remaining absorber outputs of $s$ and $r$ are unconstrained.
\end{definition}

\begin{remark}
The Lean formalization additionally requires $s \dotop (r \dotop x) = x$ on core, making $s$ and $r$ mutual inverses. This stronger condition is used only for section-side placement (Theorem~\ref{thm:placement}) and the tight cardinality bound $|S| \geq 5$ when $s \neq r$ (Theorem~\ref{thm:card}). All other results --- including the no-associativity theorem, the ICP equivalence, and the full independence theorem --- depend only on the one-sided condition $r \dotop (s \dotop x) = x$.
\end{remark}

The retraction pair provides encoding infrastructure: element $a$ can be represented as $s^k(z_1)$ for its index $k$, and $r$ peels one layer. Informally, this acts as a finite internal coding scheme, analogous in spirit to a G\"odel numbering.

\begin{definition}[Classifier Dichotomy]
\label{def:cd}
An extensional 2-pointed magma satisfies the \emph{classifier dichotomy} when:
\begin{enumerate}
  \item A \emph{classifier} $\tau \in \core(S)$ exists with $\tau \dotop x \in \{z_1, z_2\}$ for all $x$.
  \item For every $y \in \core(S)$, either
    \begin{itemize}
      \item $y \dotop x \in \{z_1, z_2\}$ for all $x \in \core(S)$ \quad ($y$ is a \emph{classifier}), or
      \item $y \dotop x \notin \{z_1, z_2\}$ for all $x \in \core(S)$ \quad ($y$ is a \emph{non-classifier}).
    \end{itemize}
  \item At least one non-classifier exists (excluding the degenerate case where the ``dichotomy'' has only one inhabited class).
\end{enumerate}
\end{definition}

The dichotomy creates a three-category decomposition $S = Z \sqcup C \sqcup N$ where $Z = \{z_1, z_2\}$ (absorbers), $C$ (classifiers), and $N$ (non-classifiers). No element can be partially classifying. This suggests a loose analogy with Kripke's grounded/ungrounded stratification of truth~\cite{kripke75}: classifier-like elements are separated from computation-producing ones. We use the comparison only as intuition, not as a formal correspondence.

\begin{definition}[Internal Composition Property]
\label{def:icp}
A magma $(S, \dotop, z_1, z_2)$ satisfies the \emph{Internal Composition Property} $(\ICP)$ when:
\[
\exists\, a, b, c \in \core(S),\ \text{pairwise distinct, such that:}
\]
\begin{enumerate}
  \item \emph{Core-preservation}: $b \dotop x \notin \{z_1, z_2\}$ for all $x \in \core(S)$.
  \item \emph{Factorization}: $a \dotop x = c \dotop (b \dotop x)$ for all $x \in \core(S)$.
  \item \emph{Non-triviality}: $|\{a \dotop x : x \in \core(S)\}| \geq 2$.
\end{enumerate}
\end{definition}

ICP says: the left-regular representation admits a non-trivial factorization on core. One core action factors through two others, one of them core-preserving — a single fragment of internal composition. In Section~\ref{sec:icp} we prove that $\ICP$ is equivalent to the standard operational axioms Compose+Inert (Theorem~\ref{thm:icp-equiv}).

\begin{definition}[Capability S]
An extensional 2-pointed magma has capability $S$ (self-representation) if it admits a retraction pair.
\end{definition}

\begin{definition}[Capability D]
An extensional 2-pointed magma has capability $D$ (self-description) if it satisfies the classifier dichotomy (Definition~\ref{def:cd}: existence of a classifier, the global dichotomy, and non-degeneracy).
\end{definition}

\begin{definition}[Capability H]
An extensional 2-pointed magma has capability $H$ (self-execution) if it satisfies the $\ICP$.
\end{definition}

All three capabilities are properties of the same base structure (Definition~\ref{def:epm}). None is defined as an extension of any other, making the independence theorem (Section~\ref{sec:independence}) a statement about six non-implications, all proved by Lean-verified counterexamples.

% ══════════════════════════════════════════════════════════════
\section{Universal Theorems}
\label{sec:universal}
% ══════════════════════════════════════════════════════════════

The following theorems are proved by pure algebraic reasoning (no \texttt{decide}) and formalized in \texttt{CatKripkeWallMinimal.lean} and \texttt{NoCommutativity.lean}. We state the hypotheses explicitly: some require only D, while others use both S and D.

\subsection{Theorems from D alone}

\begin{theorem}[Three-Category Decomposition]
\label{thm:three-cat}
In any extensional 2-pointed magma satisfying D, every element is a zero, a classifier, or a non-classifier. The classes are pairwise disjoint: in particular, $C \cap N = \emptyset$ (no element is partially classifying).
\end{theorem}

\begin{theorem}[Asymmetry]
\label{thm:asymmetry}
No extensional magma with two distinct left-absorbers is commutative.
\end{theorem}

\subsection{Theorems from S+D}

\begin{theorem}[Retraction Pair Placement]
\label{thm:placement}
In any extensional 2-pointed magma satisfying both S and D, any witness $r$ to S is a non-classifier: $r \in N$. (Under the stronger mutual-inverse condition in the Lean formalization, $s \in N$ also holds.)
\end{theorem}

\begin{theorem}[Cardinality Bounds]
\label{thm:card}
$|S| \geq 4$ for any extensional 2-pointed magma satisfying both S and D. This bound is tight: a 4-element witness with $s = r$ is exhibited in Appendix~\ref{app:tables}. Under the stronger mutual-inverse condition, $|S| \geq 5$ when $s \neq r$ (also tight).
\end{theorem}

\begin{theorem}[No Right Identity]
\label{thm:no-rid}
No extensional 2-pointed magma satisfying both S and D has a right identity element.
\end{theorem}

\begin{theorem}[No Associativity]
\label{thm:no-assoc}
No extensional 2-pointed magma carrying both a classifier and a retraction pair is associative. In particular, no extensional 2-pointed magma satisfying both S and D is a semigroup, monoid, or group.
\end{theorem}

\begin{proof}
Assume $(\tau \dotop a) \dotop b = \tau \dotop (a \dotop b)$ for all $a, b$. We derive a contradiction in four steps. The proof uses only extensionality, the existence of a classifier $\tau$, and a retraction pair $(s, r)$ — neither the global dichotomy nor the non-classifier existence axiom is needed. Lean file: \texttt{CatKripkeWallMinimal.lean}.

\emph{Step 1: $\tau$ absorbs on the right.} For any $a, b$:
\[
\tau \dotop (a \dotop b)
= (\tau \dotop a) \dotop b
= z_i \dotop b
= z_i
= \tau \dotop a
\]
where the second equality uses $\tau \dotop a \in \{z_1, z_2\}$ (classifier), and the third uses left-absorption. So $\tau \dotop (a \dotop b) = \tau \dotop a$ for all $a, b$.

\emph{Step 2: $\tau$ is constant on core.} For any core $x$: the retraction pair gives $x = r \dotop (s \dotop x)$. Applying Step~1 with $a = r$, $b = s \dotop x$:
\[
\tau \dotop x = \tau \dotop (r \dotop (s \dotop x)) = \tau \dotop r.
\]
Set $v = \tau \dotop r \in \{z_1, z_2\}$. Then $\tau \dotop x = v$ for all core $x$.

\emph{Step 3: $\tau$ is constant on absorbers.} Since $\tau$ is a classifier, it hits both $z_1$ and $z_2$ — otherwise, if $\tau \dotop x = z_1$ for all $x$, then $\tau$ and $z_1$ have identical rows, so $\tau = z_1$ by extensionality, contradicting $\tau \neq z_1$ (and symmetrically for $z_2$). Let $b_1, b_2$ witness $\tau \dotop b_1 = z_1$ and $\tau \dotop b_2 = z_2$. Applying Step~1 with $a = \tau$:
\[
\tau \dotop z_1 = \tau \dotop (\tau \dotop b_1) = \tau \dotop \tau = v,
\qquad
\tau \dotop z_2 = \tau \dotop (\tau \dotop b_2) = \tau \dotop \tau = v.
\]
The middle equalities use Step~1; $\tau \dotop \tau = v$ because $\tau$ is core (Step~2).

\emph{Step 4: Contradiction.} Steps 2 and 3 give $\tau \dotop x = v$ for all $x \in S$. Since $v \in \{z_1, z_2\}$, the classifier $\tau$ has the same row as absorber $v$. By extensionality, $\tau = v \in \{z_1, z_2\}$, contradicting $\tau \neq z_1, z_2$.
\end{proof}

\subsection{Theorems from S alone}

The following results show that capability S (self-representation) alone already constrains the algebra. They hold for any $\FRM$ that \emph{self-simulates} — i.e., there exists a term $t$ in the free term algebra over $S$ such that evaluating $t$ on the encodings of any $a, b$ returns the atom $a \dotop b$. In an algebra with S, the retraction pair provides encoding infrastructure; self-simulation is the process of using it to compute the Cayley table. Proved in \texttt{SelfSimulation.lean}.

\begin{theorem}[Partial Application Injectivity]
\label{thm:injectivity}
If a self-representing $\FRM$ self-simulates, the partial application map $a \mapsto \text{eval}(\text{App}(t, \text{rep}(a)))$ is injective.
\end{theorem}

\begin{corollary}
The encoding $\text{rep} : S \to \text{Term}(S)$ is injective, and equal partial applications imply identical rows.
\end{corollary}

% ══════════════════════════════════════════════════════════════
\section{Decomposition Invariance}
\label{sec:invariance}
% ══════════════════════════════════════════════════════════════

\begin{definition}[Absorber-Preserving Isomorphism]
An \emph{absorber-preserving isomorphism} $\varphi : M_1 \to M_2$ between extensional 2-pointed magmas is a bijection $\varphi : \Fin(n) \to \Fin(n)$ satisfying:
\begin{itemize}
  \item $\varphi(a \dotop_1 b) = \varphi(a) \dotop_2 \varphi(b)$ for all $a, b$
  \item $\varphi(z_1^{(1)}) = z_1^{(2)}$ and $\varphi(z_2^{(1)}) = z_2^{(2)}$
\end{itemize}
\end{definition}

\begin{theorem}[Functoriality]
\label{thm:functoriality}
Any absorber-preserving isomorphism between extensional 2-pointed magmas satisfying D preserves the three-category decomposition: $\varphi$ maps zeros to zeros, classifiers to classifiers, and non-classifiers to non-classifiers.
\end{theorem}

\begin{proof}[Proof sketch]
\emph{Zeros}: if $a$ is a left-absorber, then $\varphi(a) \dotop_2 \varphi(x) = \varphi(a \dotop_1 x) = \varphi(a)$ for all $x$ in the image of $\varphi$; by surjectivity, $\varphi(a)$ is a left-absorber.

\emph{Classifiers}: if $y \dotop_1 x \in \{z_1, z_2\}$ for all core $x$, then $\varphi(y) \dotop_2 \varphi(x) = \varphi(y \dotop_1 x) \in \{z_1^{(2)}, z_2^{(2)}\}$ for all core $\varphi(x)$; by surjectivity and injectivity (to exclude absorber preimages), $\varphi(y)$ is a classifier.

\emph{Non-classifiers}: symmetric argument using injectivity to transport $\notin \{z_1, z_2\}$.

The full proof is algebraic (no \texttt{decide}) and formalized in \texttt{Functoriality.lean}.
\end{proof}

% ══════════════════════════════════════════════════════════════
\section{ICP Characterization}
\label{sec:icp}
% ══════════════════════════════════════════════════════════════

An algebra with S can encode, decode, and look up any entry in the Cayley table via the retraction pair $(s, r)$. But this requires an \emph{external} machine to supply control flow: dispatch, composition, recursion. H internalizes a non-trivial composition pattern that would otherwise be supplied externally: dedicated elements whose left-multiplication implements composition ($\eta = \rho \circ g$) and storage ($g$). The $\ICP$ captures exactly this: one core action factors through two others, meaning the algebra contains a fragment of its own control flow as an element.

The standard formulation uses two operational axioms:
\begin{itemize}
  \item \emph{Compose}: $\eta \dotop x = \rho \dotop (g \dotop x)$ for all $x \in \core(S)$
  \item \emph{Inert}: $g \dotop x \notin \{z_1, z_2\}$ for all $x \in \core(S)$
\end{itemize}
with $\eta, g, \rho$ pairwise distinct non-absorbers and $\eta$ non-trivial ($\geq 2$ distinct values on core).

\begin{theorem}[$\ICP$ Equivalence]
\label{thm:icp-equiv}
For any finite magma on a 2-pointed set $(S, \dotop, z_1, z_2)$ of any size $n$:
\[
\ICP \iff \text{Compose} + \text{Inert} \text{ (with non-trivial composite)}.
\]
\end{theorem}

\begin{proof}
The bijection between witness triples is $(\eta, g, \rho) = (a, b, c)$: the composed element $a$ is $\eta$, the core-preserving element $b$ is $g$, and the outer element $c$ is $\rho$. Under this identification the factorization $a \dotop x = c \dotop (b \dotop x)$ becomes $\eta \dotop x = \rho \dotop (g \dotop x)$ (Compose), core-preservation of $b$ becomes Inert, and the remaining conditions (pairwise distinct, non-absorber) transfer directly.

The pairwise-distinctness and non-triviality conditions are not cosmetic — they block two degeneracies, and both are proved necessary by Lean-verified witnesses (\texttt{ICP.lean}). Without pairwise-distinctness, the retraction pair $(s, r)$ with $s = r$ trivially satisfies factorization via $s \dotop x = s \dotop (s \dotop x)$, giving every $\FRM$ a spurious $\ICP$; the Kripke-4 model (the minimal S{+}D model; Cayley table in Appendix~\ref{app:tables}) separates the weakened definition from the full one (theorem \texttt{icp\_distinct\_necessary}). Without non-triviality, any constant-on-core element $a$ (with $a \dotop x = v$ for all core $x$) trivially factors through any core-preserving $b$ by taking $c$ with $c \dotop y = v$; a custom 6-element $\FRM$ separates the weakened definition from the full one (theorem \texttt{icp\_nontrivial\_necessary}, \texttt{ICP.lean}).

The proof is pure equational logic — no \texttt{decide}, no case analysis on $n$. It holds for every finite magma on a 2-pointed set, of any size. Lean file: \texttt{ICP.lean}, theorem \texttt{icp\_iff\_composeInert}.
\end{proof}

This equivalence gives H a single-concept definition: the left-regular representation contains a non-trivially composed element. The operational axioms (Compose, Inert) and the $\ICP$ are the same property under variable renaming. Additionally, $\ICP$ is invariant under $\FRM$ isomorphisms (algebraic proof, no \texttt{decide}; \texttt{ICP.lean}, theorem \texttt{FRMIso.preserves\_icp}).

\paragraph{Enrichments.}
Two additional axioms act as \emph{bridges} between capabilities, not capabilities themselves:
\begin{itemize}
  \item \emph{Branch}: $\rho \dotop x = f \dotop x$ if $\tau \dotop x = z_1$, else $\rho \dotop x = g \dotop x$. Branch connects D to H: the classifier's judgment controls the evaluator's dispatch. It requires both D and H as prerequisites.
  \item \emph{Y} (fixed-point): $Y \dotop \rho = \rho \dotop (Y \dotop \rho)$ with $Y \dotop \rho \in \core(S)$. Y extends H across the decidability boundary: bounded evaluation becomes unbounded recursion. It requires H as a prerequisite.
\end{itemize}
Both are used in the accompanying artifact (a compiled reflective tower) but are not required for the independence or invariance results of this paper. Each is irredundant in the sense that removing it from the full axiom stack (S+D+H+Branch+Y) causes the reflective tower to fail; this was verified by SAT.

\begin{figure}[t]
\centering
\begin{tabular}{@{}r@{~}l@{\quad}l@{}}
\textsc{Input}: & $\text{rep}(a),\ \text{rep}(b)$ & \emph{S provides: $s, z_1$} \\
$\to$ \textsc{Decode}: & peel $s$-layers via $r$ & \emph{S provides: $r$} \\
$\to$ \textsc{Lookup}: & $\text{table}[a][b]$ & \emph{the algebra itself} \\
$\to$ \textsc{Dispatch}: & branch on classifier & \emph{D provides: $\tau$} \\
$\to$ \textsc{Compose}: & $\eta \dotop x = \rho \dotop (g \dotop x)$ & \emph{H provides: $\eta, g, \rho$} \\
$\to$ \textsc{Encode}: & $\text{rep}(\text{result})$ & \emph{S provides: $s, z_1$} \\
\textsc{Output}: & $\text{rep}(a \dotop b)$ & \\
\end{tabular}
\caption{Information flow in self-simulation (when all three capabilities co-occur). S provides encoding/decoding (I/O). The algebra provides the lookup (computation). D organizes elements into roles (classification). H internalizes a non-trivial composition pattern (evaluation). Without H, an external machine must supply the Compose and Dispatch steps.}
\label{fig:flow}
\end{figure}

% ══════════════════════════════════════════════════════════════
\section{Independence}
\label{sec:independence}
% ══════════════════════════════════════════════════════════════

\begin{figure}[t]
\centering
\setlength{\unitlength}{1em}
\begin{picture}(14,8)
% Nodes
\put(7,0.5){\makebox(0,0){\framebox{S}}}
\put(2,6.5){\makebox(0,0){\framebox{D}}}
\put(12,6.5){\makebox(0,0){\framebox{H}}}
% Crossed-out downward arrows S ⇏ D, S ⇏ H
\put(6.2,1.3){\vector(-1,1){3.4}}
\put(4.5,3.5){\makebox(0,0){\small$\times$}}
\put(7.8,1.3){\vector(1,1){3.4}}
\put(9.5,3.5){\makebox(0,0){\small$\times$}}
% Crossed-out upward arrows D ⇏ S, H ⇏ S
\put(3.6,5.5){\vector(1,-1){2.6}}
\put(4.9,4.2){\makebox(0,0){\small$\times$}}
\put(10.4,5.5){\vector(-1,-1){2.6}}
\put(9.1,4.2){\makebox(0,0){\small$\times$}}
% Crossed-out horizontal arrows D ⇏ H, H ⇏ D
\put(3.5,6.9){\vector(1,0){7}}
\put(7,6.9){\makebox(0,0){\small$\times$}}
\put(10.5,6.1){\vector(-1,0){7}}
\put(7,6.1){\makebox(0,0){\small$\times$}}
% Labels
\put(3.3,3.1){\makebox(0,0){\scriptsize$N\!=\!8$}}
\put(10.7,3.1){\makebox(0,0){\scriptsize$N\!=\!6$}}
\put(5.7,4.8){\makebox(0,0){\scriptsize$N\!=\!5$}}
\put(8.3,4.8){\makebox(0,0){\scriptsize$N\!=\!6$}}
\put(5.2,7.5){\makebox(0,0){\scriptsize$N\!=\!10$}}
\put(8.8,7.5){\makebox(0,0){\scriptsize$N\!=\!10$}}
\end{picture}
\caption{Full independence structure. All six pairwise non-implications are proved by Lean-verified finite counterexamples at the indicated sizes. No capability implies any other.}
\label{fig:independence}
\end{figure}

\begin{theorem}[Full Independence]
\label{thm:independence}
No capability implies any other. Specifically, all six pairwise non-implications hold:
\begin{enumerate}
  \item $S \not\Rightarrow D$: An extensional 2-pointed magma of size $8$ satisfying S contains a classifier element but fails the global dichotomy condition.
  \item $S \not\Rightarrow H$: An extensional 2-pointed magma of size $6$ satisfying S has 4 core elements (enough for $\ICP$ to be non-vacuous) but no triple satisfies the $\ICP$.
  \item $D \not\Rightarrow H$: An extensional 2-pointed magma of size $10$ satisfying both S and D has no triple satisfying the $\ICP$.
  \item $H \not\Rightarrow D$: An extensional 2-pointed magma of size $10$ satisfies $\ICP$ but violates the classifier dichotomy.
  \item $D \not\Rightarrow S$: An extensional 2-pointed magma of size $5$ satisfies the classifier dichotomy but admits no retraction pair.
  \item $H \not\Rightarrow S$: An extensional 2-pointed magma of size $6$ satisfies $\ICP$ but admits no retraction pair.
\end{enumerate}
\end{theorem}

\begin{proof}
Each direction is proved by exhibiting a concrete Cayley table and verifying the claimed properties by \texttt{native\_decide} (or \texttt{decide} for $N \leq 8$). The tables are:

\paragraph{$S \not\Rightarrow D$ ($N\!=\!8$, the Countermodel).}
An $8 \times 8$ Cayley table (Appendix~\ref{app:tables}) defining an $\FRM$ with $s = 2$, $r = 3$. The retraction pair works: $r \dotop (s \dotop x) = x$ for all core $x \in \{2,\ldots,7\}$ (e.g., $3 \dotop (2 \dotop 5) = 3 \dotop 6 = 5$). Element~4 is a classifier: its row on core is $(0,0,0,0,1,0) \subseteq \{z_1, z_2\}$. But element~5's row on core is $(6, 2, 1, 1, 1, 1)$ — it maps core inputs 2 and 3 to non-boolean values (6 and 2), while mapping inputs 4--7 to boolean value~1. This mixed behavior violates the classifier dichotomy: element~5 is neither entirely boolean-valued nor entirely non-boolean-valued on core. Self-representation holds; self-description fails. Lean file: \texttt{Countermodel.lean}.

\paragraph{$S \not\Rightarrow H$ ($N\!=\!6$, structural).}
A $6 \times 6$ Cayley table (Appendix~\ref{app:tables}) defining an extensional 2-pointed magma with a retraction pair ($s = r = 2$, an involution on core) but no $\ICP$. The core has 4 elements — more than the 3 pairwise distinct witnesses $\ICP$ requires — yet no triple satisfies the $\ICP$. Lean proves $\neg\,\texttt{HasICP}$ by exhaustive search via \texttt{decide}. Lean file: \texttt{E2PM.lean}, theorem \texttt{s\_not\_implies\_icp\_structural}.

\paragraph{$D \not\Rightarrow H$ ($N\!=\!10$).}
A $10 \times 10$ Cayley table satisfying D (and also S). Lean proves that no triple $(a, b, c)$ of pairwise distinct non-absorbers satisfies the $\ICP$, by exhaustive search via \texttt{native\_decide}. Lean files: \texttt{Countermodels10.lean}, \texttt{ICP.lean} (theorem \texttt{dNotH\_no\_icp}).

\paragraph{$H \not\Rightarrow D$ ($N\!=\!10$).}
A $10 \times 10$ Cayley table defining an extensional 2-pointed magma with a retraction pair, satisfying $\ICP$ (witnessed by $a\!=\!8, b\!=\!6, c\!=\!7$), yet violating the dichotomy: element~5 has both boolean and non-boolean outputs on core. Lean file: \texttt{Countermodels10.lean}.

\paragraph{$D \not\Rightarrow S$ ($N\!=\!5$).}
A $5 \times 5$ Cayley table defining an extensional 2-pointed magma with the classifier dichotomy (classifier at index~2, non-classifiers at 3 and 4), but admitting no retraction pair. Lean proves $\neg\,\texttt{HasRetractPair}$ by exhaustive search over all candidate pairs via \texttt{decide}. Lean file: \texttt{E2PM.lean}, theorem \texttt{d\_not\_implies\_s}.

\paragraph{$H \not\Rightarrow S$ ($N\!=\!6$).}
A $6 \times 6$ Cayley table defining an extensional 2-pointed magma satisfying $\ICP$ (witnessed by $a\!=\!2, b\!=\!4, c\!=\!5$), but admitting no retraction pair. Lean file: \texttt{E2PM.lean}, theorem \texttt{h\_not\_implies\_s}.

All tables are generated by Z3 SAT solver and independently verified. The Cayley tables are frozen in the repository for reproducibility.
\end{proof}

\paragraph{Methodology.}
Each counterexample was found by encoding the relevant capability axioms (and the negation of the target capability) as a SAT problem over an $n \times n$ Cayley table, then solving with Z3. The search space is $n^{n^2}$ (all possible binary operations on $\Fin(n)$); the axioms reduce this to a manageable constraint system. The sizes $N\!=\!8$ and $N\!=\!10$ were discovered by incrementing $n$ from the minimum until SAT returned a model. Once found, each table was frozen in \texttt{counterexamples.json} and independently verified in Lean. The counterexample sizes are minimal in the sense that SAT returned UNSAT at all smaller sizes with the same role assignments, but we have not proved universal minimality.

\paragraph{Interpretation.}
The $H \not\Rightarrow D$ direction shows that an algebra can internalize a non-trivial composition pattern, as captured by the $\ICP$, without also satisfying the classifier dichotomy. Thus self-execution in our sense does not force the organizational separation expressed by D.

% ══════════════════════════════════════════════════════════════
\section{Coexistence}
\label{sec:bound}
% ══════════════════════════════════════════════════════════════

\begin{theorem}[Coexistence Witness]
\label{thm:bound}
The three capabilities S, D, and H are simultaneously satisfiable. A Lean-verified witness exists at $N\!=\!6$.
\end{theorem}

\begin{proof}
A concrete $6 \times 6$ Cayley table (Appendix~\ref{app:tables}) is exhibited that satisfies S (retraction pair at $s\!=\!2, r\!=\!3$), D (classifier at $\tau\!=\!4$, dichotomy verified), and H ($\ICP$ witnessed by $a\!=\!2, b\!=\!3, c\!=\!5$). The retraction pair members serve simultaneously as $\ICP$ witnesses, demonstrating that role overlap enables coexistence at small sizes. Lean file: \texttt{Witness6.lean}, theorem \texttt{sdh\_witness\_6}.
\end{proof}

\begin{proposition}[Separated-Role Witness]
\label{thm:separated}
There is a witness at $N\!=\!10$ in which the principal roles (2 absorbers, 2 retraction pair, 1 classifier, 3 $\ICP$ witnesses, 2 enrichment elements) are pairwise distinct.
\end{proposition}

\begin{proof}
A concrete $10 \times 10$ Cayley table (Appendix~\ref{app:tables}) satisfies all three capabilities with 10 distinct role assignments. Additionally, the enrichments Branch and Y hold. Lean file: \texttt{Witness10.lean}.
\end{proof}

\paragraph{Minimum cardinality.}
Whether $N\!=\!5$ admits S+D+H is open. SAT returns UNSAT at $N\!=\!5$ with the retraction pair fixed at $(s, r) = (2, 3)$ iterating over all possible classifier assignments among the remaining core elements, but this does not rule out other role assignments or overlapping role encodings. We have not proved a universal lower bound.

% ══════════════════════════════════════════════════════════════
\section{Categorical Context}
\label{sec:categorical}
% ══════════════════════════════════════════════════════════════

Each capability connects to standard categorical vocabulary.

\paragraph{S = section-retraction pair.} Definition~\ref{def:frm} is a section-retraction pair $(s, r)$ with $r \circ s = \mathrm{id}$ on core~\cite{maclane71} in the endomorphism category of the magma. The Lean formalization uses the stronger mutual-inverse condition ($s \circ r = \mathrm{id}$ on core as well), but no main theorem depends on it (see the remark following Definition~\ref{def:frm}).

\paragraph{D $\approx$ decidable subobject classifier.} The classifier dichotomy parallels the $\Omega$-valued / non-$\Omega$-valued partition in a topos~\cite{johnstone02}. The three-category decomposition and its functoriality (Theorem~\ref{thm:functoriality}) are proved; the identification of $C$ with a subobject classifier image is an analogy (our totality condition — no partial classifiers — has no direct topos-theoretic counterpart).

\paragraph{H $\approx$ partial internal composition.} The $\ICP$ asks that the left-regular representation $\{L_a\}$ contains one non-trivially composed element $L_a = L_c \circ L_b$. This loosely echoes closure under composition or internal-hom structure in richer categorical settings, but is much weaker: full closure under composition fails in all tested models. The $\ICP$ is proved equivalent to Compose+Inert (Theorem~\ref{thm:icp-equiv}) and invariant under isomorphisms, but whether it corresponds to a standard categorical universal property is open.

% ══════════════════════════════════════════════════════════════
\section{Related Work}
\label{sec:related}
% ══════════════════════════════════════════════════════════════

\paragraph{Reflective towers.}
Smith's 3-Lisp~\cite{smith84} introduced the reflective tower — an infinite chain of meta-interpreters. Subsequent work~\cite{friedman84,danvy88} explored reification, intensions, and live meta-modification in reflective towers. Amin and Rompf~\cite{amin18} showed how to compile through the tower via stage polymorphism. All of these systems combine self-representation, self-description, and self-execution without separating them. Our contribution is the separation and the proof that it is genuine (no capability implies any other).

\paragraph{Self-interpretation.}
Mogensen~\cite{mogensen92} constructs self-interpreters for the $\lambda$-calculus. Brown and Palsberg~\cite{brown16} show that self-interpretation is possible in strongly-normalizing typed languages, contradicting earlier impossibility results. These works study self-interpretation (our S) but do not address the independence of S from D or H.

\paragraph{Algebraic effects and handlers.}
Plotkin and Power~\cite{plotkin03} decompose computational effects into algebraic operations and generic effects. Plotkin and Pretnar~\cite{plotkin09} show that effect handlers are algebra homomorphisms from a free model; there is a structural resemblance to section-retraction patterns familiar from algebraic semantics, although our setting is much more elementary: a finite single-sorted algebra with distinguished elements rather than free/forgetful constructions over effect signatures. Hyland, Levy, Plotkin, and Power~\cite{hyland06} show that the continuation monad is not algebraic; the flavor of obstruction is similar to our no-associativity result (Theorem~\ref{thm:no-assoc}), though the mechanisms differ — failure of a monad to admit a presentation by operations and equations vs.\ an algebraic impossibility in a finite magma. The algebraic effects literature studies finitary theories over infinite carriers; our contribution operates in the complementary setting of finite carriers with no equations.

\paragraph{Finite model theory.}
Independence results via finite counterexamples are standard in finite model theory~\cite{libkin04}. Our contribution is not the technique but the specific separation: three independently-motivated capabilities of reflective computation, shown independent in an algebraic setting, with machine-checked proofs.

% ══════════════════════════════════════════════════════════════
\section{Discussion}
\label{sec:discussion}
% ══════════════════════════════════════════════════════════════

\paragraph{The classifier dichotomy is organizational.}
In an algebra with S, the retraction pair provides the encoding to compute the Cayley table. D organizes that computation into coherent categories — classifiers that test and non-classifiers that transform. The $H \not\Rightarrow D$ result (Section~\ref{sec:independence}) shows that an algebra can internalize a non-trivial composition pattern without organizing its elements into stable semantic roles. Informally, the separation between S and D marks a difference between having an internal coding mechanism and having a clean internal role structure. The classifier dichotomy is an organizational property rather than a compositional one in the sense captured by H: it contributes semantic structure rather than forcing the compositional behavior captured by the $\ICP$.

\paragraph{Why magmas?}
Non-associativity is essential for D in our setting. Theorem~\ref{thm:no-assoc} shows that associativity forces the classifier to have a constant row, collapsing it to an absorber: no associative extensional 2-pointed magma satisfies the classifier structure required for self-description.

\paragraph{Scope and limitations.}
Whether the three-capability separation appears in other settings ($\lambda$-calculus, typed systems, infinite algebras) is the primary open question. Theorem~\ref{thm:no-assoc} constrains the target: any such translation must work without associativity or identity. A categorical characterization of H as a universal property (stronger than the analogy in Section~\ref{sec:categorical}) is also open.

\paragraph{The cost of legibility.}
The three capabilities coexist at $N\!=\!6$ (Section~\ref{sec:bound}), and the enrichments Branch and Y also hold in that witness. Additional organizational axioms push the minimum to $N\!=\!12$ (SAT-verified) and produce interpretable role structure; these are the subject of a companion paper.

\paragraph{Realization freedom.}
Each capability admits multiple axiom forms. For H, replacing $\eta = \rho \circ g$ with alternatives (e.g., $g \circ \rho$) produces different classifier distributions; the independence result holds for all variants.

% ══════════════════════════════════════════════════════════════
\section{Conclusion}
% ══════════════════════════════════════════════════════════════

We have shown that three capabilities of reflective computation — self-representation, self-description, and self-execution — are pairwise independent over finite extensional 2-pointed magmas. The six non-implications are machine-checked in Lean~4, the decomposition induced by D is invariant under isomorphism, and associativity is incompatible with the classifier structure used by D (Theorem~\ref{thm:no-assoc}), ruling out semigroups, monoids, and groups. We also exhibit concrete algebras in which all three capabilities coexist, including a Lean-verified witness at $N\!=\!6$.

\paragraph{Open problems.}
\begin{enumerate}
  \item \emph{Cross-formalism universality}: Does the three-capability separation appear in the $\lambda$-calculus, typed settings, or infinite algebras? The connection to algebraic effects (Section~\ref{sec:related}) suggests the concepts translate, but the no-associativity obstruction (Theorem~\ref{thm:no-assoc}) constrains the target setting.
  \item \emph{Categorical characterization of H}: Is the ICP equivalent to a standard universal property (e.g., existence of a partial internal hom)?
  \item \emph{Minimum coexistence bound}: The smallest known S+D+H witness has $N\!=\!6$ (Lean-verified; Section~\ref{sec:bound}). A universal lower bound is open.
\end{enumerate}

\paragraph{Artifact.}
All Lean files, counterexample tables, SAT reproduction scripts, and the 16-element reflective tower artifact are available at \texttt{[repository URL]}.

% ══════════════════════════════════════════════════════════════
% References
% ══════════════════════════════════════════════════════════════

\begin{thebibliography}{10}

\bibitem{smith84}
Brian Cantwell Smith.
\newblock Reflection and semantics in {Lisp}.
\newblock In \emph{POPL}, pages 23--35, 1984.

\bibitem{friedman84}
Daniel P. Friedman and Mitchell Wand.
\newblock Reification: Reflection without metaphysics.
\newblock In \emph{LFP}, pages 348--355, 1984.

\bibitem{danvy88}
Olivier Danvy and Karoline Malmkj{\ae}r.
\newblock Intensions and extensions in a reflective tower.
\newblock In \emph{LFP}, pages 28--39, 1988.

\bibitem{amin18}
Nada Amin and Tiark Rompf.
\newblock Collapsing towers of interpreters.
\newblock In \emph{POPL}, Article 52, pages 1--33, 2018.

\bibitem{mogensen92}
Torben Mogensen.
\newblock Efficient self-interpretation in lambda calculus.
\newblock \emph{Journal of Functional Programming}, 2(3):279--291, 1992.

\bibitem{brown16}
Matt Brown and Jens Palsberg.
\newblock Breaking through the normalization barrier: A self-interpreter for {F-omega}.
\newblock In \emph{POPL}, pages 5--17, 2016.

\bibitem{maclane71}
Saunders Mac~Lane.
\newblock \emph{Categories for the Working Mathematician}.
\newblock Springer, 1971.

\bibitem{johnstone02}
Peter~T. Johnstone.
\newblock \emph{Sketches of an Elephant: A Topos Theory Compendium}, vol.~1.
\newblock Oxford University Press, 2002.

\bibitem{libkin04}
Leonid Libkin.
\newblock \emph{Elements of Finite Model Theory}.
\newblock Springer, 2004.

\bibitem{burris81}
Stanley Burris and H.~P. Sankappanavar.
\newblock \emph{A Course in Universal Algebra}.
\newblock Springer, 1981.

\bibitem{kripke75}
Saul Kripke.
\newblock Outline of a theory of truth.
\newblock \emph{Journal of Philosophy}, 72(19):690--716, 1975.

\bibitem{hyland06}
Martin Hyland, Paul Blain Levy, Gordon Plotkin, and John Power.
\newblock Combining effects: Sum and tensor.
\newblock \emph{Theoretical Computer Science}, 357(1--3):70--99, 2006.

\bibitem{plotkin03}
Gordon Plotkin and John Power.
\newblock Algebraic operations and generic effects.
\newblock \emph{Applied Categorical Structures}, 11(1):69--94, 2003.

\bibitem{plotkin09}
Gordon Plotkin and Matija Pretnar.
\newblock Handlers of algebraic effects.
\newblock In \emph{ESOP}, pages 80--94, 2009.

\bibitem{moura21}
Leonardo de~Moura and Sebastian Ullrich.
\newblock The {Lean}~4 theorem prover and programming language.
\newblock In \emph{CADE}, pages 625--635, 2021.

\end{thebibliography}

% ══════════════════════════════════════════════════════════════
\appendix
% ══════════════════════════════════════════════════════════════

\section{Cayley Tables}
\label{app:tables}

All tables are extracted from the Lean source files and verified by \texttt{lake build}. Element $i$'s row is row~$i$; entry $(i,j)$ is $i \dotop j$.

\paragraph{$N\!=\!4$: Minimal S+D witness ($s = r$).}
Roles: $0 = z_1$, $1 = z_2$, $2 = \tau$, $3 = s = r$. \quad Lean: \texttt{CatKripkeWallMinimal.lean}, \texttt{kripke4}.

\medskip
{\small
\begin{tabular}{c|cccc}
$\dotop$ & 0 & 1 & 2 & 3 \\\hline
0 & 0 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 & 1 \\
2 & 0 & 1 & 0 & 1 \\
3 & 0 & 0 & 2 & 3 \\
\end{tabular}}

\paragraph{$N\!=\!5$: Minimal S+D witness ($s \neq r$).}
Roles: $0 = z_1$, $1 = z_2$, $2 = s$, $3 = r$, $4 = \tau$. \quad Lean: \texttt{CatKripkeWallMinimal.lean}, \texttt{kripke5}.

\medskip
{\small
\begin{tabular}{c|ccccc}
$\dotop$ & 0 & 1 & 2 & 3 & 4 \\\hline
0 & 0 & 0 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 & 1 & 1 \\
2 & 1 & 0 & 3 & 4 & 2 \\
3 & 0 & 2 & 4 & 2 & 3 \\
4 & 0 & 1 & 1 & 0 & 0 \\
\end{tabular}}

\paragraph{$N\!=\!6$: S+D+H Coexistence (role overlap).}
Roles: $0 = z_1$, $1 = z_2$, $2 = s$ (also ICP $a$), $3 = r$ (also ICP $b$), $4 = \tau$, $5 =$ ICP $c$. \quad Lean: \texttt{Witness6.lean}.

\medskip
{\small
\begin{tabular}{c|cccccc}
$\dotop$ & 0 & 1 & 2 & 3 & 4 & 5 \\\hline
0 & 0 & 0 & 0 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
2 & 3 & 3 & 4 & 2 & 5 & 3 \\
3 & 0 & 1 & 3 & 5 & 2 & 4 \\
4 & 0 & 0 & 1 & 0 & 1 & 1 \\
5 & 2 & 2 & 5 & 4 & 3 & 2 \\
\end{tabular}}

\paragraph{$N\!=\!8$: $S \not\Rightarrow D$ Counterexample.}
Roles: $0 = z_1$, $1 = z_2$, $2 = s$, $3 = r$, $4 =$ classifier, $5, 7 =$ mixed (violate dichotomy). \quad Lean: \texttt{Countermodel.lean}.

\medskip
{\small
\begin{tabular}{c|cccccccc}
$\dotop$ & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 \\\hline
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
2 & 3 & 3 & 7 & 3 & 4 & 6 & 5 & 2 \\
3 & 0 & 1 & 7 & 3 & 4 & 6 & 5 & 2 \\
4 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
5 & 6 & 2 & 6 & 2 & 1 & 1 & 1 & 1 \\
6 & 0 & 0 & 5 & 2 & 2 & 2 & 2 & 2 \\
7 & 2 & 2 & 2 & 1 & 2 & 2 & 6 & 3 \\
\end{tabular}}

\paragraph{$N\!=\!6$: $S \not\Rightarrow H$ Counterexample (structural).}
Roles: $0 = z_1$, $1 = z_2$, $2 = s = r$ (involution on core). Retraction pair holds; no triple satisfies $\ICP$ despite 4 core elements. \quad Lean: \texttt{E2PM.lean} (\texttt{sNoH6\_e2pm}).

\medskip
{\small
\begin{tabular}{c|cccccc}
$\dotop$ & 0 & 1 & 2 & 3 & 4 & 5 \\\hline
0 & 0 & 0 & 0 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
2 & 0 & 3 & 3 & 2 & 5 & 4 \\
3 & 2 & 4 & 5 & 5 & 1 & 4 \\
4 & 5 & 3 & 0 & 0 & 3 & 2 \\
5 & 4 & 2 & 2 & 2 & 2 & 2 \\
\end{tabular}}

\paragraph{$N\!=\!10$: $D \not\Rightarrow H$ Counterexample.}
Roles: $0 = z_1$, $1 = z_2$, $2 = s$, $3 = r$, $4 = \tau$. No triple satisfies $\ICP$. \quad Lean: \texttt{Countermodels10.lean} (\texttt{dNotH}), \texttt{ICP.lean} (\texttt{dNotH\_no\_icp}).

\medskip
{\footnotesize
\begin{tabular}{c|cccccccccc}
$\dotop$ & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\\hline
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
2 & 3 & 3 & 2 & 3 & 4 & 5 & 6 & 7 & 9 & 8 \\
3 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 9 & 8 \\
4 & 0 & 0 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
5 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
6 & 2 & 3 & 9 & 9 & 9 & 9 & 9 & 9 & 9 & 8 \\
7 & 3 & 2 & 9 & 9 & 9 & 9 & 9 & 9 & 9 & 8 \\
8 & 1 & 0 & 1 & 0 & 1 & 1 & 1 & 1 & 0 & 0 \\
9 & 0 & 1 & 0 & 1 & 0 & 1 & 1 & 0 & 1 & 1 \\
\end{tabular}}

\paragraph{$N\!=\!10$: $H \not\Rightarrow D$ Counterexample.}
Roles: $0 = z_1$, $1 = z_2$, $2 = s$, $3 = r$, $4 = \tau$. $\ICP$ holds ($a\!=\!8, b\!=\!6, c\!=\!7$). Element~5 violates dichotomy. \quad Lean: \texttt{Countermodels10.lean} (\texttt{hNotD}).

\medskip
{\footnotesize
\begin{tabular}{c|cccccccccc}
$\dotop$ & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\\hline
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
2 & 3 & 1 & 3 & 4 & 9 & 6 & 8 & 5 & 7 & 2 \\
3 & 0 & 1 & 9 & 2 & 3 & 7 & 5 & 8 & 6 & 4 \\
4 & 0 & 0 & 1 & 1 & 1 & 1 & 1 & 1 & 0 & 0 \\
5 & 0 & 0 & 2 & 0 & 0 & 0 & 0 & 0 & 3 & 1 \\
6 & 2 & 2 & 2 & 8 & 3 & 9 & 4 & 7 & 9 & 7 \\
7 & 8 & 3 & 2 & 8 & 3 & 9 & 4 & 7 & 3 & 1 \\
8 & 9 & 2 & 2 & 3 & 8 & 1 & 3 & 7 & 1 & 7 \\
9 & 2 & 2 & 2 & 2 & 4 & 7 & 6 & 7 & 2 & 0 \\
\end{tabular}}

\paragraph{$N\!=\!5$: $D \not\Rightarrow S$ Counterexample.}
No retraction pair exists. Element~2 is a classifier: row on core is $(1,0,0) \subseteq \{z_1, z_2\}$. Elements~3 and 4 are non-classifiers: all core outputs $\geq 2$. Dichotomy holds; no pair of core elements forms a retraction pair. \quad Lean: \texttt{E2PM.lean} (\texttt{dNoS\_e2pm}).

\medskip
{\small
\begin{tabular}{c|ccccc}
$\dotop$ & 0 & 1 & 2 & 3 & 4 \\\hline
0 & 0 & 0 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 & 1 & 1 \\
2 & 0 & 0 & 1 & 0 & 0 \\
3 & 4 & 2 & 4 & 2 & 4 \\
4 & 2 & 2 & 3 & 3 & 3 \\
\end{tabular}}

\paragraph{$N\!=\!6$: $H \not\Rightarrow S$ Counterexample.}
No retraction pair exists. $\ICP$ holds: $a\!=\!2, b\!=\!4, c\!=\!5$ with $b$ core-preserving and $2 \dotop x = 5 \dotop (4 \dotop x)$ on core. \quad Lean: \texttt{E2PM.lean} (\texttt{hNoS\_e2pm}).

\medskip
{\small
\begin{tabular}{c|cccccc}
$\dotop$ & 0 & 1 & 2 & 3 & 4 & 5 \\\hline
0 & 0 & 0 & 0 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
2 & 3 & 3 & 2 & 2 & 2 & 4 \\
3 & 0 & 2 & 2 & 2 & 4 & 0 \\
4 & 4 & 3 & 2 & 2 & 4 & 5 \\
5 & 0 & 0 & 2 & 2 & 2 & 4 \\
\end{tabular}}

\paragraph{$N\!=\!10$: S+D+H Coexistence Witness (all roles distinct).}
Roles: $0 = z_1$, $1 = z_2$, $2 = s$, $3 = r$, $4 = \tau$, $5 = f$, $6 = g$ (Inert), $7 = \rho$ (Branch), $8 = \eta$ (Compose), $9 = Y$. \quad Lean: \texttt{Witness10.lean}.

\medskip
{\footnotesize
\begin{tabular}{c|cccccccccc}
$\dotop$ & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\\hline
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
2 & 3 & 3 & 4 & 3 & 7 & 5 & 9 & 6 & 8 & 2 \\
3 & 0 & 1 & 9 & 3 & 2 & 5 & 7 & 4 & 8 & 6 \\
4 & 0 & 0 & 1 & 1 & 1 & 0 & 0 & 0 & 1 & 1 \\
5 & 2 & 2 & 7 & 2 & 8 & 9 & 4 & 3 & 4 & 2 \\
6 & 0 & 0 & 6 & 4 & 8 & 7 & 3 & 3 & 4 & 9 \\
7 & 2 & 2 & 6 & 4 & 8 & 9 & 4 & 3 & 4 & 9 \\
8 & 2 & 2 & 4 & 8 & 4 & 3 & 4 & 4 & 8 & 9 \\
9 & 3 & 4 & 7 & 3 & 9 & 2 & 2 & 9 & 2 & 3 \\
\end{tabular}}

\section{Proof Inventory}
\label{app:inventory}

All Lean files compile with \texttt{lake build}, zero \texttt{sorry}. Proof styles: \emph{Algebraic} = pure equational reasoning (no \texttt{decide}); \emph{decide} = verified by kernel computation ($N \leq 8$); \emph{native\_decide} = verified by compiled native code ($N = 10$).

\medskip
{\small
\begin{tabular}{lrcl}
\textbf{File} & \textbf{Thms} & \textbf{Style} & \textbf{Content} \\
\hline
CatKripkeWallMinimal & 17 & Algebraic & Decomposition, bounds, wall \\
NoCommutativity      &  3 & Algebraic & Asymmetry \\
Functoriality        &  4 & Algebraic & Invariance under iso \\
SelfSimulation       &  5 & Algebraic & Partial app.\ injectivity \\
ICP                  & 20 & Alg.+decide & ICP $\Leftrightarrow$ Compose+Inert \\
Countermodel         &  5 & decide & $S \not\Rightarrow D$ ($N\!=\!8$) \\
Countermodels10      &  9 & native\_dec. & $D \not\Rightarrow H$, $H \not\Rightarrow D$ \\
E2PM                 & 10 & decide & $D \not\Rightarrow S$, $H \not\Rightarrow S$, $S \not\Rightarrow H$ \\
Witness10            &  6 & native\_dec. & S+D+H at $N\!=\!10$ \\
Witness6             &  3 & decide & S+D+H at $N\!=\!6$ \\
\hline
\textbf{Total}       & \textbf{82} & & \\
\end{tabular}}

\end{document}
